\chapter{\abstractname}

In-database processing of data mining algorithms enables high-performance, real-time data analysis on up-to-date data, without the need of an additional system or software. For this purpose, data mining algorithms have to be implemented directly in the database engine. In this work we present an approach that integrates data mining with HyPer, a high-performance hybrid OLTP \& OLAP main memory database system.
\\
As proof-of-concept, the k-Means clustering algorithm is implemented as a HyPer operator benefiting of efficient code and query compilation. Two serial versions of the k-Means algorithm are implemented, demonstrating the possibilities of operator implementations in HyPer: One version with a strong focus on high-level C++ constructs, and a second version based on low-level code compilation techniques. Furthermore, a parallel version shows the advantages of running data mining algorithms on multi core machines. 
\\
Extensive experiments are performed. As performance criterion the time per iteration is used to compare the different implementations in a fair manner. The evaluation demonstrates the benefits of the low-level code compilation techniques over high-level C++ code. This knowledge can be used as building block for further data mining algorithms. Furthermore, the experiments on a variety of data sets show that the HyPer k-Means operator is able to compete with existing solutions such as Julia and R, and even outnumbers them when using the parallel implementation on a multi core machine. 

